Modelling Covid's effect on Assaults in Toronto

### 2022-04-27 outputs/paper/paper.Rmd
commiting code changes

---
title: "Covid-19's effect on crime in Toronto neighbourhoods"
subtitle: ""
author: 
  "Varun Vijay"
thanks: "Code and data are available at: https://github.com/Varun1005473462/final_folder-main.git."
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "This paper presents an analysis on the effect of Covid-19 on crime rates in Toronto neighborhoods in 2020. The paper uses datasets obtained from opendatatoronto to show how covid-19 had a negative effect on crime rates. The analysis consists of using the statistical programming language R to model the data using regression and constructing various plots and graphs to visually represent the data. The results obtained show us how pandemics like covid-19 can affect a person's mindset, enabling us to better predict their actions during such situations.\\par\\textbf {Keywords:} Covid-19, Crime. Toronto neighborhoods, residents, open data toronto"

@@

# Introduction

After the first reported outbreak of covid-19 in December 2019, the entire world changed. Covid-19 was quickly declared to be a health emergency of global concern by the World health organization in March of 2020. Governments took measures to restrict the spread of the disease such as imposing lockdowns, travel bans, restrictions on public gatherings, closing down businesses and advising people to isolate at home. Citizens of countries were essentially restricted to their homes and rarely left. Covid-19 had a huge impact on the global economy and had restricted people's actions and movements. The question this paper seeks to answer is how the covid pandemic had affected crime in various neighborhoods in Toronto in 2020.

In this paper, we use data on covid cases and crime rates in Toronto for the year 2020 to see how the number of covid cases in a neighborhood had affected crime in Toronto neighborhoods. We construct a model using R in order to see how the number of covid cases can affect the crime rate in a neighborhood. We also use the dataset to construct various histograms, bar plots and summary statistics about the dataset to improve the quality of our analysis. The results obtained seem to imply that the covid pandemic had a negative effect on crime in 2020 as crime rates went down in most Toronto neighborhoods. The results obtained contribute to our understanding of how the presence of a common threat that affects everyone such as a pandemic in this case can discourage people from criminal actions. It will allow us to improve our predictions of a person's behavior during such situations.

The paper first presents an overview of the datasets and the variables we will be using for this study in the data section. The model section presents the regression steps taken to generate the model that shows how covid affects crime. The results section discusses the various conclusions we can draw from the data and the model and finally the discussion section discusses the strenghts and weaknesses of the paper. Any additional details are included in the appendix.

# Data

## Obtaining the covid_cases dataset

```{r dataextract, echo=FALSE}
options(readr.show_col_types = FALSE)
covid_cases<-
  read_csv(here::here("inputs/data/covid_cases.csv"))
covid_cases
```

## Obtaining the neighborhood crime cases dataset

```{r dataextract2, echo=FALSE}
neighbourhood_crime_rates<-
  read_csv(here::here("inputs/data/neighbourhood_crime_rates.csv"))
neighbourhood_crime_rates
```

## Finding the number of covid cases for each neighborhood for the year 2020

```{r, echo=TRUE}
covid_cases<-
  covid_cases |>
  mutate(year=format(covid_cases$`Episode Date`,"%Y"))
covid_cases2020<-
  covid_cases |>
  filter(year == 2020)
covid_cases_neighbourhood<-data.frame(table(covid_cases2020['Neighbourhood Name']))
colnames(covid_cases_neighbourhood)[1]<-"Neighbourhood"
colnames(covid_cases_neighbourhood)[2]<-"Number of Covid cases"
covid_cases_neighbourhood 
```
## Arranging the neighborhoods in the neighborhodd_crime_rates and covid_cases_neighborhood dataset alphabetically.

```{r}
neighbourhood_crime_rates |>
  arrange(Neighbourhood)
covid_cases_neighbourhood |>
  arrange(Neighbourhood)
```

@@

\begin{equation}
Pr(\theta | y) = \frac{Pr(y | \theta) Pr(\theta)}{Pr(y)}  (\#eq:bayes)
\end{equation}

@@

@@

>>>>

---
title: "Modelling Covid's effect on Assaults in Toronto"
subtitle: "A statistical analysis of how the rise of Covid in 2020 affected assault rates in various downtown neighborhoods in Toronto"
author: 
  "Varun Vijay"
thanks: "Code and data are available at: https://github.com/Varun1005473462/final_folder-main.git."
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "This paper presents an analysis on the effect of Covid on assault rates in downtown Toronto neighbourhoods in 2020. The paper uses datasets obtained from opendatatoronto to show how Covid had a negative effect on crime rates. The analysis consists of using the statistical programming language R to model the data using regression and constructing various plots and graphs to visually represent the data. THe results obtained will allow us to better predict people's behavior. With the better prediction of people's behaviour, the police can better manage the public order and government will be able to make reasonable and effective policy to reduce the crime rate.\\par\\textbf {Keywords:} Covid, Assaults, Crime,Toronto neighbourhoods, residents, open data toronto"

@@

# 1 Introduction

After the first reported outbreak of Covid in December 2019, the entire world changed. Covid was quickly declared to be a health emergency of global concern by the World health organization in March of 2020. Governments took measures to restrict the spread of the disease such as imposing lockdowns, travel bans, restrictions on public gatherings, closing down businesses and advising people to isolate at home. Citizens of countries were essentially restricted to their homes and rarely left. Covid had a huge impact on the global economy and had restricted people's actions and movements. However, this paper wants to analyse how Covid has affected one of the most common types of crime, assaults. Assaults are often physical and in a time where physical contact is discouraged, it seems contradictory to assault someone as there is the risk you might catch covid. The question this paper seeks to answer is how the Covid pandemic had affected assault rates in various neighbourhoods in Toronto for the year 2020.

In this paper, we use data on Covid cases and crime statistics in Toronto neighbourhoods for the year 2020 to see how the number of Covid cases in a neighbourhood had affected assualts in that neighbourhood and what factors affected the assault rate the most. We constructed a model using R in order to see how Covid can affect the assault rate in a neighborhood. We also use the dataset to construct various histograms, bar plots and summary statistics about the dataset to improve the quality of our analysis. The results obtained seem to imply that the Covid pandemic had a negative effect on assaults in 2020 as crime rates went down in most downtown neighborhoods. The results obtained contribute to our understanding of how the presence of a common threat that affects everyone such as a pandemic in this case can discourage people from criminal actions. It will allow us to improve our predictions of a person's behavior during such situations. Better predictions of people's behaviors will allow the police to better manage the public order and government to make reasonable and effective policy to reduce the crime rate

The paper first presents an overview of the datasets and the variables we will be using for this study in the data section. The model section presents the regression steps taken to generate the model that shows how Covid affects crime. The results section discusses the various conclusions we can draw from the data and the model and finally the discussion section discusses the strengths and weaknesses of the paper. Any additional details are included in the appendix.

# 2 Data

## 2.1 Data Source and collection

For the dataset, we obtain data using opendatatoronto on crime related statistics, Covid related data pertaining to Toronto neighborhoods.

## 2.2 Data Modification

## 2.3 Data Tables

```{r table 1, echo=FALSE}
options(readr.show_col_types = FALSE)
covid_cases<-
  read_csv(here::here("inputs/data/covid_cases.csv"))
covid_cases<-
  covid_cases |>
  mutate(year=format(covid_cases$`Episode Date`,"%Y"))
covid_cases2020<-
  covid_cases |>
  filter(year == 2020)
covid_cases_gender<-covid_cases2020 %>% group_by(`Neighbourhood Name`, `Client Gender`) %>% count()
covid_cases_gender<-
  covid_cases_gender |>
  pivot_wider(names_from = `Client Gender`,values_fill =0,values_from = n)
covid_cases_outbreak<-covid_cases2020 %>% group_by(`Neighbourhood Name`,`Outbreak Associated`) %>% count()
covid_cases_outbreak<-
  covid_cases_outbreak |>
  pivot_wider(names_from = `Outbreak Associated`,values_fill =0,values_from = n)
covid_cases_final<-merge(covid_cases_gender,covid_cases_outbreak)
covid_cases_final<-
  covid_cases_final |>
  mutate(total_cases=`Outbreak Associated`+Sporadic)
colnames(covid_cases_final)[1]<-"Neighbourhood"
colnames(covid_cases_final)[2]<-"FEMALE_Covid_Cases"
colnames(covid_cases_final)[3]<-"MALE_Covid_Cases"
covid_cases_final<-na.omit(covid_cases_final)
```

```{r table 2, echo=FALSE}
neighbourhood_crime_rates<-
  read_csv(here::here("inputs/data/neighbourhood_crime_rates.csv"))
neighbourhood_crime_rates_assaults<- 
  neighbourhood_crime_rates |>
  select(Neighbourhood,
         Assault_2019,
         Assault_2020)
neighbourhood_crime_rates_assaults<-arrange(neighbourhood_crime_rates_assaults,Neighbourhood)
```

## Obtaining the final dataset.

```{r table 3, echo=FALSE, fig.cap="Final Dataset"}
covid_cases_final<-arrange(covid_cases_final,Neighbourhood)
neighbourhood_crime_rates_assaults<-arrange(neighbourhood_crime_rates_assaults,Neighbourhood)
final_dataset<-
  neighbourhood_crime_rates_assaults |>
  mutate(FEMALE_Covid_CASES=covid_cases_final$FEMALE_Covid_Cases,MALE_Covid_CASES=covid_cases_final$MALE_Covid_Cases,transgender_covid_cases=covid_cases_final$TRANSGENDER,unknown_covid_cases= covid_cases_final$UNKNOWN,other_covid_cases=covid_cases_final$OTHER, Outbreak_Associated=covid_cases_final$`Outbreak Associated`,Sporadic=covid_cases_final$Sporadic,Total_CASES=covid_cases_final$total_cases)
final_dataset
```

## 2.4 Data Visualization

```{r summary 1, echo=FALSE}
final_dataset |>
  datasummary_skim()
```

```{r summary 2,echo=FALSE}
final_dataset |>
  datasummary_correlation()
```

```{r figure 1, echo=FALSE, fig.cap="Assualts in 2020 compared to assaults in 2019"}
colors<-c("Assaults 2019"="red", "Assaults 2020"="blue")
ggplot()+
geom_point(data=final_dataset, mapping=aes(x=Neighbourhood, y=Assault_2019,color="Assaults 2019"),size=1) + 
geom_point(data=final_dataset, aes(x=Neighbourhood, y=Assault_2020,color="Assaults 2020"),size=1)+
  labs(x="Neighbourhoods",
       y="Assaults 2019, Assaults 2020",
       title="Assualts in 2020 compared to assaults in 2019")
```

```{r figure 2, echo=FALSE, fig.cap="Covid cases for each neighborhood in 2020"}
colors<-c("Female"="red", "Male"="blue", "Transgender"="purple","Unknown"="green","Other"="yellow")
ggplot()+
geom_point(data=final_dataset, mapping=aes(x=Neighbourhood, y=FEMALE_Covid_CASES,color="Female"),size=1) + 
geom_point(data=final_dataset, aes(x=Neighbourhood, y=MALE_Covid_CASES,color="Male"),size=1)+
  geom_point(data=final_dataset, aes(x=Neighbourhood, y=transgender_covid_cases,color="Transgender"),size=1)+
  geom_point(data=final_dataset, aes(x=Neighbourhood, y=unknown_covid_cases,color="Unknown"),size=1)+
  geom_point(data=final_dataset, aes(x=Neighbourhood, y=other_covid_cases,color="Other"),size=1)+
  labs(x="Neighbourhoods",
       y="Female, Male, Transgender, Unknown, Other",
       title="Covid cases by gender and neighborhood")
```

@@

$$ y = \beta_0 +\beta_1*x_1+\beta_2*x_2  $$

y= number of assaults in a neighbourhood
x_1 = number of female covid cases in a neighbourhood
x_2 = number of male covid cases in a neighbourhood

@@

## Table of estimates

```{r model estimates, echo=FALSE}
Terms<-c("Intercept", "Female_covid_cases estimate", "Male_covid_cases estimate", "Number of observation", "R^2", "R^2 adjusted", "AIC", "BIC", "Log.Lik", "F")
values<-c(58.410, -1.174, 2.029, 140, 0.363, 0.354, 1698.6, 1710.3, -845.280, 39.044)
data.frame(Terms, values)
```

$$ y = 58.410	 -1.174*x_1+2.029*x_2  $$

```{r figure , echo=FALSE, fig.cap="Relationship between Covid and Assault rates in 2020"}
final_dataset |>
  ggplot(mapping = aes(x = Total_CASES, y = Assault_2020,color="")) +
  geom_point()
```

```{r, echo=FALSE}
final_dataset |>
  ggplot(aes(x = FEMALE_Covid_CASES+MALE_Covid_CASES, y = Assault_2020)) +
  geom_point() +
  geom_smooth(method = "lm",
              color = "black", 
              linetype = "dashed",
              formula = 'y ~ x') +
  labs(x = "FEMALE_Covid_CASES+MALE_Covid_CASES",
       y = "total_Crime_2020") +
  theme_classic() +
  scale_color_brewer(palette = "Set1")
```

```{r}
plot(final_model)
```

@@

## AIC selection
```{r, echo=FALSE}
first_model <-
  lm(Assault_2020 ~FEMALE_Covid_CASES,
     data = final_dataset)
second_model <-
  lm(Assault_2020 ~ MALE_Covid_CASES,
     data = final_dataset)
third_model <-
  lm(Assault_2020 ~FEMALE_Covid_CASES+MALE_Covid_CASES,
     data = final_dataset)
fourth_model<-
  lm(Assault_2020 ~FEMALE_Covid_CASES+MALE_Covid_CASES+Sporadic,
     data = final_dataset)
fifth_model <-
  lm(Assault_2020 ~ FEMALE_Covid_CASES+MALE_Covid_CASES+Outbreak_Associated,
     data = final_dataset)
sixth_model <-
  lm(Assault_2020 ~ FEMALE_Covid_CASES+MALE_Covid_CASES+Sporadic+Outbreak_Associated,data = final_dataset)
seventh_model<-
   lm(Assault_2020 ~ Sporadic+Outbreak_Associated,data = final_dataset)
eight_model<-lm(Assault_2020 ~ 1,data = final_dataset)
ninth_model<-lm(Assault_2020~Total_CASES,data=final_dataset)
modelsummary(list(first_model,second_model,third_model,fourth_model,fifth_model,sixth_model, seventh_model,eight_model,ninth_model))
```

```{r model, echo=FALSE, message=FALSE, include=FALSE}
final_model<-lm(formula=Assault_2020~FEMALE_Covid_CASES+MALE_Covid_CASES,data=final_dataset)
modelsummary(final_model)
```
